Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-11 paper/main.tex
move logic and syntax to the appendix

\sectskip
\section{Syntax and Semantics}
\label{sec:sem}
\aftersectskip

\iffull{
\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^1 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost')
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma
\\ \cost' = \cost {-} M_c^2 {-} M_e(e)
}
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost')
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_s)
}
%
\and
\Rule{S:Tick}
{ }
{ (\state, \code{tick}(n), \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_t(n))
}

\end{mathpar}
\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
\label{fig:opsem}
\end{figure}
}{}

\ifshort{
\begin{figure}[t!]
\small
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_s)
}
\end{mathpar}
\caption{\iffull{Rules}{Selected rules} of the operational semantics of statements.}
\label{fig:opsem}
\end{figure}
}{}

@@

\iffull{
\begin{figure*}
\begin{mathpar}
%
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
\!\!  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q + M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state + P - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  \!\! P \models P' \!\! \\
  \!\! \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \!\! \\
  \!\! Q' \models Q \!\! \\
  \!\! B' \models B \!\! \\
  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \Qplusz
}
{ \!\! \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x } \!\!
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

\sectskip
\section{Quantitative Hoare Logic}
\label{sec:logic}
\aftersectskip

\iffull{In the rule {\sc L:Skip}, we do not have to account for any resource
consumption.  As a result, the precondition $Q$ can be any (potential)
function and we only have to make sure that we do not end up with more
potential.  Since the execution of \code{skip} leaves the program
state unchanged, we can simply use the precondition as postcondition.
The potential functions $B$ for the \code{break} and $R$ for the
\code{return} part of the postcondition are not reachable and can
therefore be arbitrary.}{}

In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
be arbitrary since it is unreachable.  Instead, we have to justify the
potential functions $B$ and $R$ that hold after a \code{return} and a
\code{break}, respectively.  In {\sc L:Break}, we require to have
potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
cost of \code{break} and $B$ to pay for the potential after the
\code{break}.  In {\sc L:Return} we only require to have potential
$R$ in the precondition to pay for the potential after the
\code{return}.  The reason is that we found it to be more convenient
to account for the execution cost $M_r$ of the return in rule {\sc
  L:Call}\iffull{ for function calls.}{.}

\iffull{The rule {\sc L:Update} is the standard assignment rule of Hoare
logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
the precondition we ensure that $Q$ evaluates to the same number as in
the postcondition.  We also require that have we the constant
potential $M_u + M_e(e)$ available in the precondition to pay for the
cost of the evaluation of $e$ and the update.

}
{
The rules {\sc L:Update} and {\sc L:Seq} correspond to the respective
rules of standard Hoare logic.
}
In the rule {\sc L:Loop}, the \code{break} part of the
loop body $S$ becomes the postcondition of the loop statement. We use
an arbitrary $B$ as the \code{break} part of the judgement for
$\code{loop}\, S$ since its operational semantics ensures that it can
only terminate with a \code{skip} or a \code{return}.  The
precondition $I$ of the loop is the loop invariant.  \iffull{That is why we
require to have potential $I$ available in the precondition of the
loop body $S$.}{}  In the postcondition of $S$, the potential must be
sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
iteration.

\iffull{
The {\sc L:If} is similar to the rule for the conditional in classic
Hoare logic.  In the preconditions of the judgments for the two
branches $S_1$ and $S_2$ we lift the boolean assertions
$\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
quantitative assertions.  In the precondition of the rule {\sc L:Tick}
we account for the execution cost $M_t(n)$ of the \code{tick}
statement that depends on the integer $n$.
}{}

The rule {\sc L:Call} accounts for the execution cost of both $M_f$
for function calls and $M_r$ for \code{return} statements.
\iffull{This justifies that the {\sc L:Return} rule does not account
  for resource consumption.}{} The pre- and postcondition $P_f$ and
$Q_f$ are taken from the function context $\Delta$.  The assertions in
the context are parametric with respect to both the values of the
function arguments and the return value. This allows us to specify a
bound for a function whose resource consumption depends on its
arguments.  The arguments are instantiated by the call rule
using the result of the evaluation of the argument variables in the
current state.  \iffull{Recall that we require that $P_f$ and $Q_f$ do
  not depend on the local variables on the call stack, that is,
  $\forall z \, \vec v \, \theta \, \theta' \, \gamma . \, P_f \, z \,
  \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
  (\theta',\gamma)$.}{} To transfer potential that depends on local
variables of the callee from the precondition $P$ to the postcondition
$Q$, we use an assertion $A : \Assn$ that is independent of global
variables, that is, $\forall \theta \, \gamma \, \gamma' . \, A
(\theta,\gamma) = A(\theta,\gamma')$.  It is still possible to express
relations between global and local variables using logical
variables\iffull{ (see the following paragraph for details).}{.}

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\models$ that point-wise applies $\ge$.  \iffull{This
rule indeed weakens the statement since it requires more resource to
run the statement and yields less than what has been proved to be
available after its termination.}{}

\ifx\fullversion\undefined{}\else{
\paragraph{Logical Variables and Shallow Embedding.}

When specifying a program, it is often necessary to relate certain
invariants in the pre- and postcondition.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  Because of its flexibility, we used the latter
approach in our development.  This choice makes it possible to have
logical variables almost for free: we can simply use the variable and
binding mechanisms of Coq, our host theory.  When an logical variable
is needed we introduce it using a universal quantifier of Coq before
using the logic to derive triples.  For example, the Coq theorem for
the above example would look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function body $S_f$ with this
induction hypothesis.  In this process it can be the case that we have
to use the induction hypothesis with a different value of a logical
variable (e.g., because the values of the arguments in the recursive
call differ from the values of the arguments of the callee).  To cope
with this problem, assumptions in the function context $\Delta$ are
universally quantified over logical variables.  The {\sc L:Extend}
rule uses the host proof assistant to require that the triple on $f$'s
body is proved for every possible logical variable $y$.
}\fi

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
\begin{align*}
  (m {\le} n \land P(\state) {\le} \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
  \implies \cost' {\ge} 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used for
induction in the soundness proof for the function-call and loop cases.
The constraint $\cost' \geq 0$ imposed by the definition ensures that
the program does not stop because of a resource error (recall that a
negative resource counter on the right-hand side is a resource
failure).  However, it does not rule out memory safety errors or
assertion failures. This is because our logic does not prove any
safety or correctness theorems but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition\iffull{ of validity.}{.}

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  \iffull{We define it as follows.
\begin{align*}
\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
& & \land& \safe n Q {\code{skip}} \cont
\end{align*}}
{
We define $\safeK n B R Q \cont \mathop{:=}
\safe n B {\code{break}} \cont \, \land \,
\safe n Q {\code{skip}} \cont \, \land \,
\safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont$.

}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.

Using the semantic validity of triples we define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$. \iffull{A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.}{}  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
We are now able to state the soundness of the quantitative logic.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
  $
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two constraints force
$\delta$ to be exactly one in the theorem statement.

Assume that $S$ is a complete program and $\Delta$ is empty.  By
expanding the definitions we see that $\Delta$ is valid
for every $n$ and that $\Kstop$ is safe for every $n$. So
we derive
$$
\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
$$
This means that from any starting state $\state$, $P(\state)$
provides enough resources for any run of the program $S$.  \iffull{This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.}{}

@@

>>>>

@@

@@

\section{Quantitative Hoare Logic}
\label{app:logic}

\begin{figure*}
\begin{mathpar}
%
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule[leftskip=.0cm,rightskip=.0cm]{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + M_e(e) + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule[leftskip=.0cm,rightskip=.0cm]{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
\!\!  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q + M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state + P - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P + M_e(e) }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  \!\! P \models P' \!\! \\
  \!\! \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \!\! \\
  \!\! Q' \models Q \!\! \\
  \!\! B' \models B \!\! \\
  \!\! \forall v.\, (R'\,v \models R\,v) \!\!
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \Qplusz
}
{ \!\! \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x } \!\!
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

\caption{Rules of the Quantitative Hoare Logic}
\label{fig:logic}
\end{figure*}

In the rule {\sc L:Skip}, we do not have to account for any resource
consumption.  As a result, the precondition $Q$ can be any (potential)
function and we only have to make sure that we do not end up with more
potential.  Since the execution of \code{skip} leaves the program
state unchanged, we can simply use the precondition as postcondition.
The potential functions $B$ for the \code{break} and $R$ for the
\code{return} part of the postcondition are not reachable and can
therefore be arbitrary.

In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
be arbitrary since it is unreachable.  Instead, we have to justify the
potential functions $B$ and $R$ that hold after a \code{return} and a
\code{break}, respectively.  In {\sc L:Break}, we require to have
potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
cost of \code{break} and $B$ to pay for the potential after the
\code{break}.  In {\sc L:Return} we only require to have potential
$R$ in the precondition to pay for the potential after the
\code{return}.  The reason is that we found it to be more convenient
to account for the execution cost $M_r$ of the return in rule {\sc
  L:Call} for function calls.

The rule {\sc L:Update} is the standard assignment rule of Hoare
logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
the precondition we ensure that $Q$ evaluates to the same number as in
the postcondition.  We also require that have we the constant
potential $M_u + M_e(e)$ available in the precondition to pay for the
cost of the evaluation of $e$ and the update.

In the rule {\sc L:Loop}, the \code{break} part of the
loop body $S$ becomes the postcondition of the loop statement. We use
an arbitrary $B$ as the \code{break} part of the judgement for
$\code{loop}\, S$ since its operational semantics ensures that it can
only terminate with a \code{skip} or a \code{return}.  The
precondition $I$ of the loop is the loop invariant.  That is why we
require to have potential $I$ available in the precondition of the
loop body $S$.  In the postcondition of $S$, the potential must be
sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
iteration.

The {\sc L:If} is similar to the rule for the conditional in classic
Hoare logic.  In the preconditions of the judgments for the two
branches $S_1$ and $S_2$ we lift the boolean assertions
$\code{isfalse}~\sem{e}_\state$ and $\code{isfalse}~\sem{e}_\state$ to
quantitative assertions.  In the precondition of the rule {\sc L:Tick}
we account for the execution cost $M_t(n)$ of the \code{tick}
statement that depends on the integer $n$.

The rule {\sc L:Call} accounts for the execution cost of both $M_f$
for function calls and $M_r$ for \code{return} statements.
This justifies that the {\sc L:Return} rule does not account
  for resource consumption. The pre- and postcondition $P_f$ and
$Q_f$ are taken from the function context $\Delta$.  The assertions in
the context are parametric with respect to both the values of the
function arguments and the return value. This allows us to specify a
bound for a function whose resource consumption depends on its
arguments.  The arguments are instantiated by the call rule
using the result of the evaluation of the argument variables in the
current state.  Recall that we require that $P_f$ and $Q_f$ do
  not depend on the local variables on the call stack, that is,
  $\forall z \, \vec v \, \theta \, \theta' \, \gamma . \, P_f \, z \,
  \vec v \, (\theta,\gamma) = P_f \, z \, \vec v \,
  (\theta',\gamma)$. To transfer potential that depends on local
variables of the callee from the precondition $P$ to the postcondition
$Q$, we use an assertion $A : \Assn$ that is independent of global
variables, that is, $\forall \theta \, \gamma \, \gamma' . \, A
(\theta,\gamma) = A(\theta,\gamma')$.  It is still possible to express
relations between global and local variables using logical
variables (see the following paragraph for details).

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\models$ that point-wise applies $\ge$.  This
rule indeed weakens the statement since it requires more resource to
run the statement and yields less than what has been proved to be
available after its termination.

\paragraph{Logical Variables and Shallow Embedding.}

When specifying a program, it is often necessary to relate certain
invariants in the pre- and postcondition.  A standard solution of Hoare
logic is to use \emph{logical variables}~\cite{Kleymann99}.  These
additional variables (also called auxiliary state in the literature)
are constant across the derivation.  For example, if $Z$ is such
a logical variable, we can specify the function $\code{double}()$
which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$
%
When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  Because of its flexibility, we used the latter
approach in our development.  This choice makes it possible to have
logical variables almost for free: we can simply use the variable and
binding mechanisms of Coq, our host theory.  When an logical variable
is needed we introduce it using a universal quantifier of Coq before
using the logic to derive triples.  For example, the Coq theorem for
the above example would look as follows.
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  qtriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}
%
However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic.  First we add a specification $\forall z\,\vec
v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)$ of $f$ to the function context
$\Delta$.  Then we proceed to prove the function body $S_f$ with this
induction hypothesis.  In this process it can be the case that we have
to use the induction hypothesis with a different value of a logical
variable (e.g., because the values of the arguments in the recursive
call differ from the values of the arguments of the callee).  To cope
with this problem, assumptions in the function context $\Delta$ are
universally quantified over logical variables.  The {\sc L:Extend}
rule uses the host proof assistant to require that the triple on $f$'s
body is proved for every possible logical variable $y$.

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
\begin{align*}
  (m {\le} n \land P(\state) {\le} \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
  \implies \cost' {\ge} 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used for
induction in the soundness proof for the function-call and loop cases.
The constraint $\cost' \geq 0$ imposed by the definition ensures that
the program does not stop because of a resource error (recall that a
negative resource counter on the right-hand side is a resource
failure).  However, it does not rule out memory safety errors or
assertion failures. This is because our logic does not prove any
safety or correctness theorems but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition of validity.

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  We define it as follows.
\begin{align*}
\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
& & \land& \safe n Q {\code{skip}} \cont
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B{+}M_b{+}x} {R{+}x} {Q{+}x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.
